---
title: "Preparation of input for stan" 
comment: "This notebook takes cleaned datasets of RDS type and processes them into the lists that stan works with"

author: "Astrid Tausen"
date: today
format: html
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
library(here)
library(tidyverse)
```

```{r}
#| label: configuration

path_intermediate <- here("data", "intermediate")
path_processed    <- here("data", "processed")

if (!dir.exists(path_processed)) {
  dir.create(path_processed, recursive = TRUE)
}

# Choose subregion for 2 level exact GPR here (will not affect other models):
target_region <- "Eastern Asia"

# Timeline for trajectories:
start_year <- 1930
end_year <- 2025
normalize_time <- TRUE # Always use TRUE!! 

model_id <- format(Sys.time(), "%Y%m%d_%H%M%S")

```

```{r}
#| label: load-intermediate-data

ari_estimates  <- readRDS(here(path_intermediate, "ari_estimates.rds"))
ihme_prev <- readRDS(here(path_intermediate, "ihme_prev.rds"))
```

```{r}
#| label: setup-timeline

# Sequence of years the trajectories are generated for
all_years <- seq(start_year, end_year)

if (normalize_time) {
  x_time <- (all_years - min(all_years)) / (max(all_years) - min(all_years))
} else {
  x_time <- all_years
}

x_time <- as.numeric(x_time)

```

**Two level model (country and UN subregion)**

In the following section, the distance matrix and the list for stan is created for the first, simple Bayesian model with a 2-level hierarchical prior on the Styblo ratio. The distance matrix is created manually with squared euclidian distance between years.

```{r}
#| label: spatial-temporal-sets-2-level

# Get only the right subregion for exact GPR
df_ari_chol <- ari_estimates |> filter(sub_region == target_region)
df_prev_chol <- ihme_prev |> filter(sub_region == target_region)

years_chol <- sort(unique(c(df_ari_chol$year, df_prev_chol$year)))
countries_chol <- sort(unique(c(df_ari_chol$iso3, df_prev_chol$iso3)))

df_ari_chol$year_id <- match(df_ari_chol$year, years_chol)
df_ari_chol$country_id <- match(df_ari_chol$iso3, countries_chol)

df_prev_chol$year_id <- match(df_prev_chol$year, years_chol)
df_prev_chol$country_id <- match(df_prev_chol$iso3, countries_chol)

# Distance matrix (years)
time_dist_sq <- as.matrix(dist(years_chol, method = "euclidean"))^2
```

```{r}
#| label: assemble-stan-list-2-level
cholesky_dat_list <- list(
  N_years = length(years_chol),
  N_countries = length(countries_chol),
  
  N_ari = nrow(df_ari_chol),
  log_ari = df_ari_chol$lari, 
  ari_se = df_ari_chol$lari_se,
  ari_year = df_ari_chol$year_id,
  ari_country = df_ari_chol$country_id,
  
  N_prev = nrow(df_prev_chol),
  log_prev = df_prev_chol$lprev,
  prev_se = df_prev_chol$lprev_se,
  prev_year = df_prev_chol$year_id,
  prev_country = df_prev_chol$country_id,
  
  distance_matrix = time_dist_sq
)

output_filename_chol <- paste0("stan_list_2_level_", model_id, ".rds")
saveRDS(cholesky_dat_list, here(path_processed, output_filename_chol))

print(paste("List for stan is saved in data/processed:", output_filename_chol))
```

**Three level model (country, UN subregion and UN region)**

For the three level models, I have used stan's function gp_exp_quad_cov(), which is why the distance matrix isn't created manually.

```{r}
#| label: build-spatial-dictionary

# Get a unique map of region -> subregion -> country
hierarchy_dict <- ihme_prev |>
  select(region, sub_region, iso3) |>
  drop_na() |>
  distinct() |>
  arrange(region, sub_region, iso3)

regions <- unique(hierarchy_dict$region)
subregions <- unique(hierarchy_dict$sub_region)
countries <- unique(hierarchy_dict$iso3)
```

```{r}
#| label: create-stan-indices

# Map every subregion to its parent region
subregion_map <- hierarchy_dict |>
  select(region, sub_region) |>
  distinct() |>
  mutate(
    region_id = match(region, regions),
    subregion_id = match(sub_region, subregions)
  ) |>
  arrange(subregion_id)

stan_subregion_to_region <- subregion_map$region_id

# Map every country to its parent subregion
country_map <- hierarchy_dict |>
  mutate(
    subregion_id = match(sub_region, subregions),
    country_id = match(iso3, countries)
  ) |>
  arrange(country_id)

stan_country_to_subregion <- country_map$subregion_id
```

```{r}
#| label: index-observations

# Index ARI
dat_ari <- ari_estimates |>
  filter(iso3 %in% countries, year %in% all_years) |>
  mutate(
    country_id = match(iso3, countries),
    year_id = match(year, all_years)
  ) |>
  drop_na(lari, lari_se) 

# Index prevalence
dat_prev <- ihme_prev |>
  filter(iso3 %in% countries, year %in% all_years) |>
  mutate(
    country_id = match(iso3, countries),
    year_id = match(year, all_years)
  ) |>
  drop_na(lprev, lprev_se)
```

```{r}
#| label: assemble-stan-list_3_level

stan_dat_list <- list(
  N_years = length(all_years),
  x_time = x_time,
  N_regions = length(regions),
  N_subregions = length(subregions),
  N_countries = length(countries),
  
  country_to_subregion = stan_country_to_subregion,
  subregion_to_region = stan_subregion_to_region,

  N_ari = nrow(dat_ari),
  log_ari = dat_ari$lari,
  ari_se = dat_ari$lari_se,             
  ari_year = dat_ari$year_id,
  ari_country = dat_ari$country_id,
  
  N_prev = nrow(dat_prev),
  log_prev = dat_prev$lprev,
  prev_se = dat_prev$lprev_se,    
  prev_year = dat_prev$year_id,
  prev_country = dat_prev$country_id
)

output_filename <- paste0("stan_list_3_level_", model_id, ".rds")
saveRDS(stan_dat_list, here(path_processed, output_filename))

# Also save a mapping file so we know which indices correspond to which strings
mapping_list <- list(
  regions = regions,
  subregions = subregions,
  countries = countries,
  years = all_years
)
saveRDS(mapping_list, here(path_processed, paste0("mapping_dict_", model_id, ".rds")))

print(paste("List for stan is saved in data/processed:", output_filename))

```
